---
title: "Data wrangling in tidyr"
author: "Kiran Favre"
date: "2022-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(tidyverse)
library(here)
library(janitor)
```

## Data wrangling

#### Read in World Bank data files

```{r}
wb_indicators <- read_csv(here("data", "wb_indicators.csv"), na = c("..", ""))
# notice how empty values are populated by '..' and by empty cells! so we use na

wb_metadata <- read_csv(here("data", "wb_indicators_metadata.csv"))
```

#### How can we make this data more tidy?

##### Put all years in one column and seriesnames in one row

-   can use pivotlonger to get years into a single column

    -   start with data, then use function and think of which columns you want to gather together.

```{r}
wb_indicators_long <- wb_indicators |> 
  pivot_longer(cols = '2001 [YR2001]': '2020 [YR2020]',
               names_to = "year",
               values_to = "indicator_value")
```

#### Separate to clean up year column

```{r}
wb_clean <- wb_indicators_long |> 
  separate(col = year, into = c("year", "year_chr"),
           sep = " ") |> 
  select(-year_chr, -'Country Code', -'Series Code') |> 
  mutate(year = as.numeric(year)) |> 
  drop_na('Series Name') |> 
  pivot_wider(names_from = 'Series Name', values_from=indicator_value)

names(wb_clean) <- c("country", "year", "access_clean_fuels_pp", "access_to_electricity_pp", "co2_emissions_kt", "fossil_fuel_cons_pct", "water_stress")


# now we want to plot co2 emission by country

# to search by row, we use filter
##look for a vector in the data
wb_subset <- wb_clean |> 
  filter(country %in% c("Algeria", "Barbados", "Bulgaria", "Chile"))

ggplot(data = wb_subset, aes(x= year, y = co2_emissions_kt, group = country)) +
  geom_line(aes(color = country)) 
  # + facet_wrap(~country) if you wanted to look at this in their own plots!
```

-   separate the title "year [yearxxx]

-   select is used for selecting rows vs filter is to select rows!

-   mutate is used to give a new column name to years that are numeric (rather than characters)

-   drop_na does complete row deletion on rows with NA in the column of interest

-   widen the variables currently trapped in 'Series Name' to be spread across multiple columns and get values that populate those columns from 'Indicator Value'

    -   making each unique input for series name its own column!

-   rename columns in new data set! (names is not a pipeable function, so have to keep it outside)

## Afternoon session

-   subsetting rows based on you conditions

    -   dpylr::filter(df=...)

        -   i want to keep or exclude rows based on whether or not they satisfy conditions I specify

### Filter to only keep exact matches/ single conditions

EXAMPLE 1: We want to keep heights greater than 180.

```{r}
ex_1 <- starwars |> 
  filter(height > 180)
```

EXAMPLE 2: We want to only keep observations where the eye color is blue.

```{r}
ex_2 <- starwars |> 
  filter(eye_color == "blue")
```

EXAMPLE 3: We want to make a subset where the home world is Naboo

```{r}
ex_3 <- starwars |> 
  filter(homeworld == "Naboo")
```

### Filter to satisfy multiple conditions

EXAMPLE 4: Create a subset from starwars where the height is greater than 180 and the home world is Tatooine

```{r}
ex_4 <- starwars |> 
  filter(height > 180 &
           homeworld == "Tatooine")
```

EXAMPLE 5: Create a subset where the hair color is brown and species is human

```{r}
ex_5 <- starwars |> 
  filter(hair_color == "brown" &
           species == "Human")
```

### Filtering 'or' statements

EXAMPLE 6: Create a subset where height \> 180 or eye color is yellow

```{r}
ex_6 <- starwars |> 
  filter(height > 180 |
           eye_color == "yellow")
```

EXAMPLE 7: Create a subset that keeps rows if home world is Endor or species is a droid

```{r}
ex_7 <- starwars |> 
  filter(homeworld == "Endor" |
           species == "Droid")
```

EXAMPLE 8/9: Create a subset where eye color is blue or brown or red

```{r}
ex_8 <- starwars |> 
  filter(eye_color == "blue" | eye_color == "brown" |
           eye_color == "red")
#can get long



ex_9 <- starwars |> 
  filter(eye_color %in% c("blue", "brown", "red"))
#this goes from row to row and sees if the input is any of the colors
## versus filter(eye_color == c("blue", "brown", "red")) will look at row 1 and look for blue, then row 2 look for brown, in that order, where what we coded will look at each row and see if blue, brown or red fits the value inside. 
```

EXAMPLE 11: Create a subset where the home world is Endor or Tatooine or Naboo AND height \<150

```{r}
ex_11 <- starwars |> 
  filter(homeworld %in% c("Endor", "Tatooine", "Naboo"),
         height < 150)
```

EXAMPLE 12/13: We want to create a subset with all species besides human

```{r}
ex_12 <- starwars |> 
  filter(species != "Human")

ex_13 <- starwars |> 
  filter(!species %in% c("Human", "Ewok",
                          "Wookie"))
```

### `Select()` or exclude columns

How to select or exclude columns in a few ways:

-   Select by name

    EXAMPLE 14/15 create a subset that only looks at certain columns

```{r}
ex_14 <- wb_clean |> 
  select(country, year, co2_emissions_kt)

ex_15 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct)
```

EXAMPLE 16: we want to exclude the column "access_electricity_pp"

```{r}
ex_16 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct & -access_to_electricity_pp)
```

EXAMPLE 17: Multiple pipe steps together

```{r}
ex_17 <- wb_clean |> 
  filter(country %in% c("Bolivia", "Chile", "Mexico")) |> 
  select(country:fossil_fuel_cons_pct)
```

EXAMPLE 18: Another way to use select is to rename or move things around

```{r}
ex_18 <- wb_clean |> 
  select(year_new = year,
         emissions = co2_emissions_kt,
         country_name = country)

#kept the 3 columns in reference, renamed them
```

EXAMPLE 19: Select only columns for country, year and water stress from wb_clean data and reorder them as year, country, water stress. Then filter to only include observations for any 3 countries of your choosing.

```{r}
ex_19 <- wb_clean |> 
  select(year, country, water_stress) |> 
  filter(country %in% c("Australia", "Aruba", "Lebanon"))
```

### Rename columns using dplyr

EXAMPLE 20/21: We want to rename only 2 columns from a data set

```{r}
##rename(new name = old name)

ex_20 <- wb_clean |> 
  rename(ffcp = fossil_fuel_cons_pct,
         ws = water_stress)

#choose 2 to rename yourself
ex_21 <- wb_clean |> 
  rename(thems_w_electricity = access_to_electricity_pp,
         thems_w_fuels = access_clean_fuels_pp)
```

### `Mutate()` :update an existing column or add a new one

If you assign a new column name, it will add a new column to the end of your data frame. If you assign an existing column name, it will overwrite that column.

EXAMPLE 22: Add a new column to wb_clean that is co2_emissions_tons

```{r}
ex_22 <- wb_clean |> 
  mutate(co2_emissions_t = co2_emissions_kt * 1000) |> 
  relocate(co2_emissions_t, .after = co2_emissions_kt)

ex_23 <- wb_clean |> 
  mutate(yay = "YAY")

#change the class of year from numeric to characteristic
ex_24 <- wb_clean |> 
  mutate(year = as.character(year))

#convert back!
ex_25 <- ex_24 |> 
  mutate(year = as.numeric(year))


#dplyr::relocate()
ex_26 <- wb_clean |> 
  relocate(water_stress, .after = year)


```

### `group_by |> summarize` : creates groups under the hood based on variables you specify

#### Summarize applies your function to EACH GROUP and returns them in a nice summary table

EXAMPLE 27: group Star Wars characters by species

```{r}
ex_27 <- starwars |> 
  filter(homeworld %in% c("Naboo", "Tatooine")) |> 
  group_by(species, homeworld) |> 
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))
```

EXAMPLE 28:

```{r}
df <- tribble(
  ~species, ~length, ~number,
  "lobster", 12, 4,
  "lobster", 14, 6,
  "lobster", 15, 2)
#this data is not tidy bc of multiple observations in one row
#lobster, that bend at 12 in, there are 4 of them. 

df_case <- df |> 
  uncount(number)

```

EXAMPLE 29: count

```{r}
# can use count function to create frequency tables
# count counts observations
# combination of group_by summarize n()

ex_29 <- starwars |> 
  count(homeworld, species)


#this will do the same thing, but we can do as many calculations as we want in this method
ex_30 <- starwars |> 
  group_by(homeworld, species) |> 
  summarize(size = n(), mean_height = mean(height, na.rm = TRUE))
```

Practice example:

-   start from wb_clean

-   only keep observations from Nicaragua, Ecuador, Peru and Chile AND for years greater than 2005

-   only keep columns for country, year and co2 emissions

-   add a new column that is co2 emissions converted to megatons

-   find the total co2 emissions in megatons still in the data set by country

```{r}
practice_ex <- wb_clean |> 
  filter(country %in% c("Nicaragua", "Ecuador", "Peru", "Chile") & year > 2005) |>
    select(country, year, co2_emissions_kt) |> 
  mutate(co2_emissions_mt = co2_emissions_kt / 1000) |>
  group_by(country) |> ##next step will be grouped by country
  summarize(total_co2_mt = sum(co2_emissions_mt, na.rm = TRUE))
  
```
